---
title: "Tooth growth analysis"
author: "Jan Herman"
date: "October 25, 2015"
header-includes:
    \usepackage{amsmath}
output: 
  pdf_document: 
    fig_caption: yes
    keep_tex: yes
---
# Overview
This short analysis concerns the dataset `ToothGrowth` -- one of the standard datasets 
shipped with R. The dataset contains information about tooth growth described by length
of odontoblasts (cells responsible for tooth growth; measured in \(\mu\)m) in 60 
guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1 
and 2 mg/day) by one of two delivery methods (orange juice or ascorbic acid).

We will perform basic exploratory analysis of the data and then tests hypotheses about
an impact of supplement types and different doses on odontoblasts lenghts.

# Exploratory analysis
Throughout the analyses, we will need three R packages -- `dplyr` for data manipulation,
`ggplot2` for figure plotting and `knitr` for making pdf output (and displaying summary
tables by the `kable()` function). We will load them now.
```{r}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(knitr)
```

Next we load the dataset, rename the supplement factor levels and do basic summary.
```{r}
data(ToothGrowth) 
ToothGrowth <- ToothGrowth %>%
    mutate(supp = factor(supp, labels = c("Orange juice", "Ascorbic acid")))
summary(ToothGrowth)
```

Then we plot the first
(and only) exploratory graph. Since throughout the analyses we don't have any
ambitions to uncover a quantitative relationship between dose and odontoblasts length,
we treat dose as an (ordered) factor, rather than as a numeric value.

```{r fig.height=3,fig.cap='Exploratory graph -- odontoblasts length by dose and supplement type'}
ggplot(data = ToothGrowth, aes(x = factor(dose), y = len)) +
    geom_point(size = 4, shape = 21) + 
    facet_grid(. ~ supp) +
    ggtitle("Odontoblasts length by dose and supplement") +
    labs(x = "Dose [mg]", 
         y = expression(paste("Odontoblasts length [", mu, "m]"))) +
    theme_bw()
```

# Summaries
In this section, we present short summaries of tooth growth dataset, possibly grouped by
dose and/or supplement type. In each table we report median, sample mean and sample standard
deviation.

```{r}
summary_dataset <- tbl_df(ToothGrowth) %>%
    summarize(median = median(len), mean = mean(len), sd = sd(len))
kable(summary_dataset, col.names = c("Median", "Mean", "Standard deviation"), 
      caption = "Overall summary")
summary_by_dose <- tbl_df(ToothGrowth) %>%
    group_by(dose) %>%
    summarize(median = median(len), mean = mean(len), sd = sd(len))
kable(summary_by_dose, col.names = c("Dose", "Median", "Mean", "Standard deviation"),
      caption = "Summary by dose")
summary_by_supp <- tbl_df(ToothGrowth) %>%
    group_by(supp) %>%
    summarize(median = median(len), mean = mean(len), sd = sd(len))
kable(summary_by_supp,
      col.names = c("Supplement type", "Median", "Mean", "Standard deviation"),
      caption = "Summary by supplement type")
summary_by_supp_and_dose <- tbl_df(ToothGrowth) %>%
    group_by(supp, dose) %>%
    summarize(median = median(len), mean = mean(len), sd = sd(len))
kable(summary_by_supp_and_dose,
      col.names = c("Supplement type", "Dose", "Median", "Mean", "Standard deviation"),
      caption = "Summary by supplement type and dose")
```

#Hypotheses testing
In all hypotheses testing, we will use the Student's \(t\)-tests, since we have only 
a small 
number of samples and cannot deduce the distribution they are drawn from.

## Odontoblasts length by supplement type
At first we will concern on comparing odontoblasts length on both types
of supplement. Denote \(\mu_\mathrm{OJ}\) the mean of odontoblasts length of pigs with 
orange juice supplement, \(\mu_\mathrm{VC}\) the mean of odontoblasts length of pigs with 
ascorbic acid supplement. Recall from previous section, that 
\(\bar{x}_\mathrm{OJ}=`r round(summary_by_supp[1,3],3)`\) and 
\(\bar{x}_\mathrm{VC}=`r round(summary_by_supp[2,3],3)`\) are the 
corresponding sample means.
The null hypothesis \(H_0\) states that \(\mu_\mathrm{OJ}=\mu_\mathrm{VC}\),
the alternative hypothesis \(H_1\) states that the (population) means are unequal.

We will use both sided student's \(t\)-test (as we don't know which supplement 
should have larger impact
on odontoblasts length), with two groups (the pigs in experiment are all different,
hence impairable), with possible unequal variances of groups and the (default)
0.95 confidence level:

```{r}
dose_hyp <- t.test(len ~ supp, data = ToothGrowth, var.equal = FALSE, paired = FALSE)
dose_hyp
```

From the verbose output we see that we cannot reject \(H_0\). The confidence interval
`dose_hyp$conf.int`, which is  \((`r round(dose_hyp$conf.int[1], 3)`,
`r round(dose_hyp$conf.int[2], 3)`)\),
does contain zero. Also note that the \(p\)-value is `r dose_hyp$p.value`, i.e. greater
than \(0.05=1-0.95\), which is the bound for the 0.95 confidence level.

## Odontoblasts length by dose
Let us denote \(\mu_{d=0.5}\), \(\mu_{d=1}\) and \(\mu_{d=2}\) the population means of
pigs by the specified dose (0.5 mg, 1 mg and 2 mg, respectively). Recall that the 
corresponding sample means are \(\bar{x}_{d=0.5}=`r round(summary_by_dose[1,3], 3)`\),
\(\bar{x}_{d=1}=`r round(summary_by_dose[2,3], 3)`\) and 
\(\bar{x}_{d=2}=`r round(summary_by_dose[3,3], 3)`\), respectively.

We feel that there is a strong positive correlation between dose and tooth growth,
let us make it clear using one-sided student's \(t\)-tests. We will make three of them,
one for each pair of doses.

Let \(H_0^1\) be the null hypothesis stating \(\mu_{d=0.5}=\mu_{d=1}\) and the alternative
\(H_1^1\) will be of the form \(\mu_{d=0.5}<\mu_{d=1}\). As before, consider the two groups
unpaired and their variances unequal. Set the confidence level to 0.99.

```{r}
supp_hyp_1 <- t.test(len ~ dose, data = filter(ToothGrowth, dose %in% c(0.5, 1)),
                     var.equal = FALSE, paired = FALSE,
                     alternative = "less", conf.level = 0.99)
```

The confidence interval (`supp_hyp_1$conf.int`) is \((-\infty,
`r round(supp_hyp_1$conf.int[2], 3)`)\) does not contain 0, so we reject \(H_0^1\)
in favor of \(H_1^1\). For curiosity, the \(p\)-value corresponding for the given test
is \(`r supp_hyp_1$p.value`\) hence we should reject the \(H_0^1\) even if we
set the confidence level to 0.9999999.

Without much more effort we can state the hypotheses and test them for the remaining 
pairs of doses. The meta-settings (two group \(t\)-test, one-sided, unequal variances,
unpaired groups, confidence level 0.99) remain the same.

Let \(H_0^2\) be the null hypothesis stating \(\mu_{d=1}=\mu_{d=2}\) and the alternative
\(H_1^2\) be of the form \(\mu_{d=1}<\mu_{d=2}\). Similarly denote by \(H_0^3\) the null
hypothesis \(\mu_{d=0.5}=\mu_{d=2}\) and the alternative
\(H_1^3\) states \(\mu_{d=0.5}<\mu_{d=2}\).

```{r}
supp_hyp_2 <- t.test(len ~ dose, data = filter(ToothGrowth, dose %in% c(1, 2)),
                     var.equal = FALSE, paired = FALSE,
                     alternative = "less", conf.level = 0.99)
supp_hyp_3 <- t.test(len ~ dose, data = filter(ToothGrowth, dose %in% c(0.5, 2)),
                     var.equal = FALSE, paired = FALSE,
                     alternative = "less", conf.level = 0.99)
```

We will reject both the null hypotheses in favor of the alternatives. The evidence are
the confidence intervals \((-\infty,
`r round(supp_hyp_2$conf.int[2], 3)`)\) for the second test and \((-\infty,
`r round(supp_hyp_3$conf.int[2], 3)`)\) for the last one.
Corresponding the \(p\)-values are \(`r supp_hyp_2$p.value`\), respectively 
\(`r supp_hyp_3$p.value`\) [sic].

#Conclusions

Concerning an impact of supplement type on the odontoblasts length we cannot reject the
hypothesis that the population means of both group are equal (at the 0.95 confidence level).
In another words at this confidence level we can't tell if the supplement type of vitamin C
has an impact on tooth growth of guinea pigs.

On the other hand, we have shown that the odontoblasts length **does** depend on the dose
of vitamin C recieved by guinea pigs (at the 0.99 confidence level and possibly
a lot higher, as is suggested by reported \(p\)-values). All of three tests addressing
this dependency showed that larger dose implies longer odontoblasts.

#Notice
The author apologizes for exceeding the maximum page size of the report. But in his opinion
splitting the report into text and supplementary part each of 3 pages would decrease
the readibility of the analysis.

Another excuse for the bad English.

The knitr source code of this analysis is available at
[github](https://github.com/hermitko/StatInf_CourseProject2).

